Evaluating Perceptual Judgements on 3D Printed Bar Charts

### 2024-01-17 jds.rmd
Updates and add refs and code for sine illusion

The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar. @fig-bar-types show the two types of grouped bar charts. 
We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

@@

## 3D Graphical Perception
Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams [@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013].

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Digital graphics has developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision.

@@

Our study is designed to replicate and expand upon the position-length experiment from Cleveland and McGill as closely as possible.
In this section, we will discuss the replication process and the design of our modified version of this experiment.

@@

Each graph presents two bars from the values given above where the participants are asked to judge the ratio of the smaller bar to the larger bar. The ratio of bars used by Cleveland and McGill were 17.8, 26.1, 38.3, 46.4 (twice), 56.2, 68.1 (twice), and 82.5 (twice).
The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice.

Each graph is presented so that there are ten bars where only two of the bars are marked for identification.
Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.
Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).

>>>>

The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar. @fig:bar-types show the two types of grouped bar charts. 
We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

@@

## 3D Graphical Perception
Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion[@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013], for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in @fig:sine-illusion). 
While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception. Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.

```{r sine-illusion}
#| cache: false
#| fig-width: 4
#| fig-height: 4
#| out-width: "30%"
#| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion. All vertical lines in figures (a) and (c) are the same length, but the lines in the middle of the curve appear to be much shorter. The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth."
#| fig-subcap:
#|    - "Sine Illusion"
#|    - "3D surface with similar features"
#|    - "Perceived vs. actual length"
#' Function to data frame
#'  
#' @param n number of values 
#' @param ell extent of vertical range
#' @param x range of horizontal values
#' @param f function
#' @param fp first derivative of function f
#' @param f2p second derivative of function f
#' @return data frame with n function values and derivatives along the x axis for a range given by x
#' @example
#' f <- function(x) 2*sin(x)
#' fp <- function(x) 2*cos(x)
#' f2p <- function(x) -2*sin(x)
#' dframe <- function.frame(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
#' require(ggplot2)
#' qplot(x, y=ystart, yend=yend, xend=x, geom="segment", data=dframe) +
#'    geom_point(aes(x,y))
function.frame <- function(n=200, ell=1, x, f, fprime, f2prime) {
  x <- seq(x[1], x[2], length=n)
  if (length(ell) != length(x)) ell <- rep(ell, length(x))
  y <- f(x)
  ystart <- y - ell/2
  yend <- y + ell/2
  # now correct for line illusion in vertical direction
  dy <- diff(range(y))
  dyl <- diff(range(c(ystart, yend)))
  # fprime and f2prime are sensitive to the aspect ratio of a plot
  # we represent it in framework of dy and dy+len
  # needs to be fixed by factor a:  
  a <- dy/dyl

  fp <- a*fprime(x)
  f2p <- a*f2prime(x)
  data.frame(x,y,ystart, yend, fp, f2p, a)
}

#' This function has to be split in parts and renamed
#' 
#' Right now this function is a convoluted mess.  We need to have separate
#' functions that 
#' (1) create a data frame for a given function,
#' its first and second derivatives and range in x
#' (this is what the function function.frame is supposed to do), 
#' (2) a trig transform function
#' (3) a quadratic approach function - better even, make (2) and (3) one function and use a parameter to decide on the method to use.
#' @example
#' f <- function(x) 2*sin(x)
#' fp <- function(x) 2*cos(x)
#' f2p <- function(x) -2*sin(x)
#' dframe <- createSine(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
#' require(ggplot2)
#' qplot(x=x, xend=x, y=y+ellx4.u, yend=y-ellx4.l, geom="segment", data=dframe, linetype=I(1)) + 
#'    theme_bw() + coord_fixed(ratio=1) + 
#'    xlab("x") + ylab("y") + 
#'    scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), minor_breaks=minor.axis.correction,
#'      labels=c("0", expression(paste(pi,"/2")), expression(pi), expression(paste("3",pi, "/2")), 
#'      expression(paste("2",pi))))

createSine <- function(n=200, len=1, f=f, fprime=fprime, f2prime=f2prime, a=0, b=2*pi) {
  #  if(getquadapprox & !is.function(f2prime)) f2prime <- function(x) -1*f(x) # for backwards compatibility
  x <- seq(a, b, length=n+2)[(2:(n+1))]
  ell <- rep(len, length=length(x))
  fx <- f(x)
  ystart <- fx - .5*ell
  yend <- fx + .5*ell

  # now correct for line illusion in vertical direction
  dy <- diff(range(fx))
  dx <- diff(range(x))
  # fprime works in framework of dx and dy, but we represent it in framework of dx and dy+len
  # needs to be fixed by factor a:  
#  a <- dy/(dy + len) 
  dyl <- diff(range(c(ystart, yend)))
  a <- dy/(dyl) 
  # ellx is based on the "trig" correction
  ellx <- ell / cos(atan(abs(a*fprime(x))))
  # ellx2 is based on linear approximation of f  
  ellx2 <- ell * sqrt(1 + a^2*fprime(x)^2)

  # make this a data frame - ggplot2 doesn't do well with floating vectors
  dframe <- data.frame(x=x, xstart=x, xend=x, y=fx, ystart=ystart, yend=yend, ell=ell, ellx = ellx, ellx2=ellx2)

  # third adjustment is based on quadratic approximation of f.
  # this needs two parts: correction above and below f(x)  
  #   if(getquadapprox & is.function(f2prime)){
  #     secseg <- do.call("rbind", lapply(dframe$x, function(i) getSecantSegment(i, dframe, f, fprime, f2prime)))
  #     dframe$ellx3.u <- secseg$sec.ell1
  #     dframe$ellx3.l <- secseg$sec.ell2
  #   }

  fp <- a*fprime(x)
  f2p <- a*f2prime(x)
  v <- 1 + fp^2
  lambdapinv <- 0.5*(sqrt(v^2-f2p*fp^2*ell) + v)    
#  lambdaminv <- -0.5*(sqrt(v^2+f2p*fp^2*ell) + v)
  lambdaminv <- 0.5*(sqrt(v^2+f2p*fp^2*ell) + v)

  dframe$ellx4.l <- 0.5*abs(lambdapinv)/sqrt(v)
  dframe$ellx4.u <- 0.5*abs(lambdaminv)/sqrt(v)
#  dframe$lambdam <- 1/lambdaminv
#  dframe$lambdap <- 1/lambdapinv
  fp <- fprime(x)
  f2p <- f2prime(x)
  v <- 1 + fp^2
#  lambda1 <- (-v + sqrt(v^2 - fp^2*f2p*ell))/(fp^2*f2p) # these two work
#  lambda2 <- (-v + sqrt(v^2 + fp^2*f2p*ell))/(fp^2*f2p) # these two work

  lambdaa <- -2*(sqrt(v^2 - f2p*fp^2*ell) + v)^-1 # identical to top
  lambdab <- 2*(sqrt(v^2 + f2p*fp^2*ell) + v)^-1

  dframe$lambdam <- lambdaa
  dframe$lambdap <- lambdab

  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdab))^-1, ellx4.u, data=dframe)
  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdaa))^-1, ellx4.l, data=dframe)
  # ggplot(aes(x, y+ell*(a*sqrt(v)*abs(lambdaa))^-1, xend=x, yend=y-ell*(a*sqrt(v)*abs(lambdab))^-1), data=dframe) + geom_segment()

  dframe
}

#' what is the function doing? - in one sentence
#' 
#' function description in a paragraph
#' @param x0 is a vector of locations for which the secant segments are supposed to be calculated
#' @param df df is a data frame
#' @param f function
#' @param fp first derivative of function f
#' @param f2p second derivative of function f
#' @return data frame consisting of .... what is the output of the function?
#' @example
#' # need an example here of how to use the function
getSecantSegment <- function(x0, df, f, fprime, f2prime){
# find the closest values of the grid df corresponding to x0
  # doesn't work right now
  ell.x0 <- sapply(x0, function(x) df$ell[which.min(x-df$x)])

  dy <- diff(range(with(df, c(y+ell.x0, y-ell.x0))))
  dx <- diff(range(df$x))
  a <- dx/dy

  #---- Actual Roots
  #   secSlope   <- -a/fprime(x0)
  #   temp       <- seq(min(df$x)-pi, max(df$x)+pi, .0001)
  #   leftend    <- temp[which.min(abs(f(temp) + ell.x1 - secSlope*(temp-x1)))]
  #   rightend   <- temp[which.min(abs(f(temp) - ell.x1 - secSlope*(temp-x1)))]
  #----

  #---- Approximation
  lambda1 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 - 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))
  lambda2 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 + 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))

  x1 <- lambda1*fprime(x0)+x0
  x2 <- lambda2*fprime(x0)+x0
  y1 <- f(x0)-lambda1
  y2 <- f(x0)-lambda2
  #----

  #---- Approximation V2 ----
  #   dx <- fprime(x0)
  #   dx1 <- 1+dx^2   
  #   ddx <- f2prime(x0)
  #   
  #   corr1 <- 1/a*(((dx1)-sqrt(dx1^2-2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
  #   corr2 <- 1/a*(((dx1)-sqrt(dx1^2+2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
  #   if(abs(dx)<.01) {
  #     dx1 <- 2
  #     corr1 <- corr2 <- 0
  #   }
  #   if(abs(dx)<.01){
  #     ddx <- 1
  #   }
  #   
  #   x1 = corr1 + x0
  #   x2 = corr2 + x0
  #   y1 = f(x1)+ell.x0
  #   y2 = f(x2)-ell.x0
  #----

  df2 <- data.frame(x=x0, y=f(x0), deriv=fprime(x0),
                    sec.xstart=x1, sec.xend = x2, 
                    sec.ystart=y1, sec.yend = y2,
                    ell.orig = 2*ell.x0)

  df2$sec.ell1 <- a*.5*ell.x0/with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
  df2$sec.ell2 <- a*.5*ell.x0/with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
  #   df2$sec.ell1 <- with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
  #   df2$sec.ell2 <- with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
  df2$type <- "Perceived Width"
  df2$a <- a
  return(df2)
}

correctx <- function(z, fprime, a=0, b=2*pi, w=1) {
  # w = 1/(shrink+1)
  const <- integrate(function(x) abs(fprime(x)), a, b)$value
  trans <- sapply(z, function(i) integrate(function(x) abs(fprime(x)), a, i)$value*(b-a)/const + a)
  # alternatively to the rowMeans, you could report back  
  # trans*(1-w) + z*w
  trans*w + z*(1-w)
}

f <- function(x) 2*sin(x)
fprime <- function(x) 2*cos(x)
f2prime <- function(x) -2*sin(x)

x <- seq(0, 2*pi, length=42)[2:41]
data <- do.call("rbind", lapply(seq(-.5, .5, 1), function(i) data.frame(x=x, y=3*sin(x), z=i)))

data.persp <- reshape2::acast(data, x~z, value.var="y")
x <- sort(unique(data$x))
y <- sort(unique(data$y))
z <- sort(unique(data$z))
dframe <- createSine(n = 50, len = 1, f=f, fprime=fprime, f2prime=f2prime)
dframe %>%
  ggplot(aes(x = xstart, xend = xend, y = ystart, yend = yend)) + geom_segment() + coord_fixed() + 
  theme_void()

persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=50) # , ltheta=0, lphi=-15

# persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=3) # , ltheta=0, lphi=-15

dframe <- createSine(n = 150, len = 1, f=f, fprime=fprime, f2prime=f2prime)
dframe$ystartcts <- dframe$ystart
dframe$yendcts <- dframe$yend
dframe[1:150,c(2, 3, 5, 6)] <- NA
dframe[(1:15)*10-5, c(2, 3)] <- dframe[(1:15)*10-5, 1] 
dframe[(1:15)*10-5, 5] <- dframe[(1:15)*10-5, 4] - .5
dframe[(1:15)*10-5, 6] <- dframe[(1:15)*10-5, 4] + .5
dframe$type <- "Vertical Width"

idx <- which(!is.na(dframe$xstart))
dframe$ell <- dframe$ell/2
dframe.1 <- getSecantSegment(dframe$xstart[idx], dframe, f, fprime, f2prime)
dframe.1$x <- dframe$x[idx]
dframe.1$y <- dframe$y[idx]
dframe.1$ystartcts <- dframe$ystartcts[idx]
dframe.1$yendcts <- dframe$yendcts[idx]
names(dframe.1) <- c("x", "y", "deriv", "xstart", "xend", "ystart", "yend", "ell", "ell.quad1", "ell.quad2", "type", "a", "ystartcts", "yendcts")
dframe.1$vangle <- with(dframe.1, atan(deriv))
dframe <- bind_rows(dframe, dframe.1)
dframe$type <- factor(dframe$type)

ggplot(aes(x=x, y=y), data=dframe, colour=I("grey50")) + 
  geom_line(color = "grey") + theme_void() + 
  geom_line(aes(y=ystartcts), colour="grey50", alpha = .25) +
  geom_line(aes(y=yendcts), colour="grey50", alpha = .25) +
  geom_segment(data=subset(dframe, !is.na(type)), 
               aes(x=xstart, xend = xend, y=ystart, yend=yend, colour=type, linetype=type), linewidth=0.8)  + 
  xlab("") + ylab("")  +
  coord_equal(ratio=1) + scale_colour_manual("", values=c("blue", "grey30")) + 
  # geom_text(aes(label=paste("theta", "%~~%", round(abs(vangle)/pi*180), "^o", sep=""), 
                # x=pmax(xstart, xend)/2+x/2+.35 , y=y-sign(vangle)*.6+.02), colour="blue",
            # data=dframe.1, parse=TRUE, hjust=.9, vjust=.5, size=3) + 
  scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), 
                     labels=c("0", expression(paste(pi,"/2")), expression(pi), 
                              expression(paste("3",pi, "/2")), expression(paste("2",pi)))) +
  scale_linetype_manual("", values=c("11", "solid")) + 
  theme(legend.key.width = unit(3, "line"), plot.margin = unit(c(0,0,0,0), "cm"), legend.position=c(.25, .2))

@@

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Digital graphics has developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

@@

Our study is designed to replicate and expand upon the position-length experiment from Cleveland & McGill as closely as possible, with the hope of being able to integrate this study with previously reported results in a consistent manner. 
In this section, we discuss the replication process and the design of our version of this experiment.

@@

Each graph presents two bars from the values given above where the participants are asked to judge the ratio of the smaller bar to the larger bar. The ratio of bars used by Cleveland and McGill were 17.8, 26.1, 38.3, 46.4 (twice), 56.2, 68.1 (twice), and 82.5 (twice).
The exact numeric comparisons were not disclosed, but the comparison values used in our study were subjected to the constraints of having the same ratio values and that no value was used more than twice. \svp{Can we show the comparison values in a table?}

Each graph is presented so that there are ten bars where only two of the bars are marked for identification, \svp{as shown in @fig:bar-types.}
Cleveland and McGill did not specify the random process for the heights of the eight other bars, so we used a scaled Beta distribution with parameters that limit excessive noise around the bars used for comparisons.
Code to reproduce the data generation process, data underlying the plots used in this study, the rendered plots and STL files, anonymized user data, and analysis code can found be at [https://github.com/TWiedRW/2023-JDS-3dcharts](https://github.com/TWiedRW/2023-JDS-3dcharts).
